---
title: "Reproducible Research: Peer Assessment 1"
author: "Radu Craioveanu"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r Setup the Options for knitr behavior for code chunks}
require(knitr)
knitr::opts_chunk$set(echo=TRUE, fig=TRUE)
```

```{r Load the file(s) and preprocess the data}

dataPath <- "." 
#download dataset if not there already
fileName <- "activity.zip" 
filePath <- paste(dataPath,fileName,sep="/") 
if (!file.exists(filePath)) {  
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivit.zip" 
  download.file(url=fileURL,destfile=filePath,method="curl") 
  #Extract the data set files from the archive
  unzip(zipfile=filePath, exdir=dataPath) 
}
#read the file now into a data frame
df <- read.csv("activity.csv")

```



## What is mean total number of steps taken per day?
```{r Run mean() while ignoring NA values}
require(dplyr)
#calculate the total number of steps taken per day 
dfday <- summarise(group_by(df,date), sum(steps))
names(dfday)[2] <- "SumOfStepsPerDay"
#remove NA
dfday <- na.omit(dfday)
#list the aray of Steps
dfday$SumOfStepsPerDay
#create a histogram of the numbers of steps taken each day
hist(dfday$SumOfStepsPerDay, xlab = "Sum of Steps per Day", ylab = "Frequency", main = "Histogram of Sum of Steps per Day")
#get the mean steps per day
mean(dfday$SumOfStepsPerDay, na.rm=TRUE)
#get the median steps per day
median(dfday$SumOfStepsPerDay, na.rm=TRUE)

```



## What is the average daily activity pattern?
```{r find and plot the daily pattern}
require(ggplot2)
avginterval <- aggregate(x=list(avgsteps = df$steps), by = list(interval = df$interval), FUN = mean, na.rm = TRUE)
ggplot(data=avginterval, aes(x=interval, y = avgsteps)) + geom_line() + xlab("5 Minute Interval") + ylab("Average of Steps per Interval") + ggtitle("Average Daily Pattern")
```


## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  We will replace the missing values with the average for that interval
```{r replacing missing values}
#calculate the number of missing rows with NA
missList <- is.na(df$steps)
#FALSE and TRUE count, TRUE being the number of missing values
table(missList)
#create a merged list from the original data and the average interval data so that we can then replace the NA with the average data
dfm <- merge(df,avginterval,by.x = "interval", by.y = "interval", all)
for(i in 1:nrow(dfm) ) { if(is.na(dfm[i,"steps"])) dfm[i,"steps"] <- dfm[i,"avgsteps"]}
#now aggregate back to daily averages
dfmday <- summarise(group_by(dfm,date), sum(steps))
#change the sum column name 
names(dfmday)[2]="SumOfStepsPerDay"
#now run the histogram
hist(dfmday$SumOfStepsPerDay, xlab = "Sum of Steps per Day", ylab = "Frequency", main = "Histogram of Sum of Steps per Day with filled NA values")
#get the mean steps per day
mean(dfmday$SumOfStepsPerDay)
#get the median steps per day
median(dfmday$SumOfStepsPerDay)
```
#The impact of replacing NA values with the average Steps per index is minimal
```{r}
#mean with NA replaced
mean(dfday$SumOfStepsPerDay)
#mean with NA removed
mean(dfmday$SumOfStepsPerDay)
#median with NA replaced
median(dfmday$SumOfStepsPerDay)
#median with NA removed
median(dfday$SumOfStepsPerDay)

```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
dfms <- dfm
mydates <- as.Date(dfms$date)
weekends <- weekdays(mydates) %in% c("Saturday", "Sunday")
dfms <- transform(dfms, week = ifelse(weekends ,'weekend', 'weekday'))
dfms$week <- as.factor(dfms$week)
```

#Graph the Average Steps per day on Weekdays and Weekends.

```{r, warning = FALSE}
#clean up the NAs
dfms <- na.omit(dfms)
dfmDay <- aggregate(dfms, list(interval = dfms$interval, week = dfms$week), FUN = mean, na.rm = TRUE)
```

```{r}
ggplot(dfmDay, aes(x = interval, y = steps)) + geom_line() + facet_wrap(~week, nrow=2) + xlab("Interval") + ylab("Number of steps") + ggtitle("Average Daily Pattern")
```
